---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

## Dependencies

- Required R packages:
* knitr 
* ggplot2 
* dplyr
* RCurl

```{r Initialize, echo=FALSE , message=FALSE , warning = FALSE}
local({r <- getOption("repos"); 
       r["CRAN"] <- "http://cran.r-project.org"; options(repos=r)})

# If knitr is not present it is installed
if ("knitr" %in% rownames(installed.packages()) == FALSE) {
  install.packages("knitr")
}
library(knitr, quietly = TRUE)

# If ggplot2 is not present it is installed
if("ggplot2" %in% rownames(installed.packages()) == FALSE) {
  install.packages("ggplot2")
}
library(ggplot2, quietly = TRUE)

# If plyr is not present it is installed
if("dplyr" %in% rownames(installed.packages()) == FALSE) {
  install.packages("dplyr")
}
library(dplyr, quietly = TRUE)

# If RCurl is not present it is installed
if("RCurl" %in% rownames(installed.packages()) == FALSE) {
  install.packages("RCurl")
}

library(scales, quietly = TRUE)

# Dataset Location URL 
Dataset.Location <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
```
## Loading and preprocessing the data

The variables included in this dataset are:

- **steps**: Number of steps taking in a 5-minute interval (missing values are coded as `NA`)
- **date**: The date on which the measurement was taken in YYYY-MM-DD format
- **interval**: Identifier for the 5-minute interval in which measurement was taken. The value in base 60 representing hours and minutes (hhmm)

The dataset is stored in a comma-separated-value (CSV) file and there are a total of 17,568 observations in this dataset.


The [orignal Dataset](`r Dataset.Location`) is downloaded, unzipped and loaded into the `activity` data.frame

```{r LoadDataset, echo=TRUE}

download.file(Dataset.Location,dest="data.zip",method="curl")
unzip(zipfile = "data.zip", overwrite=TRUE)
activity <- read.csv("activity.csv")
summary(activity)
```


## What is mean total number of steps taken per day?

Using only `completeCases` the data is aggregated by `date` having the total number of `steps` by day

```{r AggretateData, echo=TRUE}
completeCases <- activity[complete.cases(activity),]
stepsByDay <- aggregate(completeCases$steps,by=list(completeCases$date), FUN=sum)
names(stepsByDay) <- c("date","totalSteps")
```

There is a histogram of the total number of steps taken each day

```{r StepsByDayHistogram, message=FALSE, echo=FALSE}

p <- ggplot(stepsByDay, aes(x=totalSteps))
p + geom_histogram(fill="white", colour="darkblue") 

```

```{r echo=FALSE}
stepsmean <-round(mean(stepsByDay$totalSteps),2)
stepsmedian <- median(stepsByDay$totalSteps)
```

In this dataset, supressing `NA` values, the **mean** and **median** values for the `totalSteps` per day are `r stepsmean ` and `r stepsmedian` respectively.


## What is the average daily activity pattern?

The pattern will be described using a time series plot of the total steps by interval (5 minutes).

```{r ActivityPattern, echo=TRUE }
stepsByInterval <- aggregate(completeCases$steps,by=list(completeCases$interval), FUN=mean)
names(stepsByInterval) <- c("interval","averageSteps")
```


This is the time series of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)

```{r ActivityPatternTimeSerie, echo=FALSE}

# The interval is converted from a hhmm value in base 60 to the number of minutes from 00:00 and
# converted to seconds and finally added to a random data

stepsByInterval$interval <- 
    as.POSIXct.numeric( 
        (  ( ( stepsByInterval$interval %/% 100) * 60 )
         + ( stepsByInterval$interval %% 100 )
        ) * 60
       , origin = "1998-01-29 00:00:00 GMT", tz = "GMT" )

# this is the plot
q <- qplot(interval,averageSteps, data=stepsByInterval, 
           xlab = "Minute of the day", ylab = "Average of steps",
           main = " Average of steps during the day")

q + geom_line(colour = "darkblue") + 
  scale_x_datetime( breaks = "1 hour" , labels = date_format("%H:%M")) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) 
 
```

```{r MaxAverage, echo=FALSE}
# Row index for the max average steps
maxIndex <- which(stepsByInterval$averageSteps==max(stepsByInterval$averageSteps))

# max interval
maxInterval <- format(stepsByInterval$interval[maxIndex],"%H:%M")
```


The 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps is **`r maxInterval`**

## Imputing missing values

In the original dataset there are `r sum(is.na(activity$steps))` rows with `NA` values in the `steps` column.

The values for this cases are calculated, and updated, using the **mean** for the same interval in complete cases.

```{r ImputingMissingValues, echo=TRUE}

# This function is used in order to calculate the mean for an interval
intervalAvg <- function (itv) {
  mean(allCases$steps[allCases$interval == itv],na.rm=TRUE)
}

# A new data frame, with all cases is created
allCases <- activity

# steps are calculated, when missing, replacing NA by
# the mean in the same interval and not change for not NA
allCases <- within(allCases, 
                   steps <- ifelse(is.na(steps),
                                   intervalAvg(interval),
                                   steps))


```

Considering all cases and the values calculated for `NA`cases the histogram is :
```{r allStepsByDaysHistogram,   message=FALSE, echo=FALSE}
stepsByAllDays <- aggregate(allCases$steps,by=list(allCases$date), FUN=sum)
names(stepsByAllDays) <- c("date","totalSteps")

r <- ggplot(stepsByAllDays, aes(x=totalSteps))
r + geom_histogram(fill="white", colour="darkgreen") 

```

```{r echo=FALSE}
stepsmeanallcases <-round(mean(stepsByAllDays$totalSteps),2)
stepsmedianallcases <- median(stepsByAllDays$totalSteps)
```

In this new dataset, including the correction in the **steps** values ( `r sum(is.na(allCases$steps))` `NA` steps values), the **mean** and **median** values for the `totalSteps` per day are `r stepsmeanallcases ` and `r stepsmedianallcases` respectively.

## Are there differences in activity patterns between weekdays and weekends?

Two new columns are added to allCases data frame (the original activity data frame including a correction for `NA`values) :

* **weekday** contains the string with the week day name 
* **weekend** indicates if the day corresponds to a weekday or weekend (Saturday or Sunday)

```{r AddWeekend, echo=TRUE}
allCases["weekday"] <- weekdays(as.Date(allCases$date))
allCases["weekend"] <- ifelse(as.POSIXlt(allCases$date)$wday >=6,"weekend","weekday")
```

Using the **weekend** column it is possible to plot this graph representing different patterns for weekdays and weekend

```{r WeekActivityPatternTimeSerie, echo=FALSE}

# The data is aggregated by interval and weekend day (true or false)
stepsByIntervalWeekday <- aggregate(allCases$steps,by=list(allCases$interval,allCases$weekend), FUN=mean)
names(stepsByIntervalWeekday) <- c("interval","weekend","averageSteps")


# The interval is converted from a hhmm value in base 60 to the number of minutes from 00:00 and
# converted to seconds and finally added to a random data

stepsByIntervalWeekday$interval <- 
    as.POSIXct.numeric( 
        (  ( ( stepsByIntervalWeekday$interval %/% 100) * 60 )
         + ( stepsByIntervalWeekday$interval %% 100 )
        ) * 60
       , origin = "1998-01-29 00:00:00 GMT", tz = "GMT" )

# this is the plot
s <- qplot(interval,averageSteps,data=stepsByIntervalWeekday,
           xlab = "Minute of the day", ylab = "Average of steps",
           main = " Average of steps during the day")

s + geom_line(colour = "black") + 
  scale_x_datetime( breaks = "1 hour" , labels = date_format("%H:%M")) +
  facet_grid( weekend ~ .) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) 

```


